---
output:
  pdf_document: default
  html_document: default
---
**LCA(Latent Class Analysis) with variables age, admission type (elective or emergency) and morbiditiy composition of each patient**
```{r}
install.packages("RPostgreSQL")
install.packages("poLCA")

library(RPostgreSQL)
library(poLCA)
```
Establish connection to database in PostgreSQL
```{r}
#Establish connection to database in PostgreSQL
conn <- dbConnect(drv="PostgreSQL",
                  dbname="mimic",
                  host= "localhost",
                  user="postgres")
```
Query* to select necessary variables -

*Assuming create_elixhauser, create_patient_info,
*and create_elixhauser_final have been run in database already,
```{r}
# *Assuming all necessary tables have 
# *have been created in database already
# Query* to select patient cohort -

query <- "SELECT * FROM ELIXHAUSER_MATCH;"
```
Store query results

```{r}
final_table <- dbGetQuery(conn, query)
```

Factor all variables for LCA (poLCA requires factored data)
```{r}
final_table$age <- as.factor(final_table$rounded_age)

final_table$admission_type_encoded <- as.factor(final_table$admission_type_encoded)

comorbidity_columns <- c(
  "congestive_heart_failure",
  "cardiac_arrhythmias",
  "valvular_disease",
  "pulmonary_circulation",
  "peripheral_vascular",
  "hypertension",
  "paralysis",
  "other_neurological",
  "chronic_pulmonary",
  "diabetes_uncomplicated",
  "diabetes_complicated",
  "hypothyroidism",
  "renal_failure",
  "liver_disease",
  "peptic_ulcer",
  "aids",
  "lymphoma",
  "metastatic_cancer",
  "solid_tumor",
  "rheumatoid_arthritis",
  "coagulopathy",
  "obesity",
  "weight_loss",
  "fluid_electrolyte",
  "blood_loss_anemia",
  "deficiency_anemias",
  "alcohol_abuse",
  "drug_abuse",
  "psychoses",
  "depression"
)


final_table[comorbidity_columns] <- lapply(
                              final_table[comorbidity_columns], as.factor)
```

Define formula for running LCA with poLCA, cbind creates matrix 
with observed variables, and "~1" indicates 1 latent class being created
```{r}
lca_formula <- cbind(rounded_age, admission_type_encoded, congestive_heart_failure, 
                     cardiac_arrhythmias, valvular_disease, 
                     pulmonary_circulation,
                     peripheral_vascular, hypertension,
                     paralysis,
                     other_neurological,
                     chronic_pulmonary,
                     diabetes_uncomplicated,
                     diabetes_complicated, 
                     hypothyroidism,
                     renal_failure,
                     liver_disease,
                     peptic_ulcer,
                     aids,
                     lymphoma,
                     metastatic_cancer,
                     solid_tumor,
                     rheumatoid_arthritis,
                     coagulopathy,
                     obesity,
                     weight_loss,
                     fluid_electrolyte,
                     blood_loss_anemia,
                     deficiency_anemias,
                     alcohol_abuse,
                     drug_abuse,
                     psychoses,
                     depression) ~1

```

Run LCA model, we are using 7 Latent classes to model the data here
```{r}
lca_out <- poLCA(lca_formula, data=final_table, nclass=7, graph=FALSE)

dbDisconnect(conn)
```

Run chi-square test for assessing preferred heterogeneity of morbidities within classes (categorical variables)
```{r}
latent_class_memberships <- lca_out$predclass
significant_p_values <- TRUE 

# Loops through each variable (morbidity) and perform chi-square tests
for (variable in comorbidity_columns) {
  contingency_table <- table(final_table[[variable]], latent_class_memberships)
  
  # Checks if there are cells with expected frequency < 5
  if (any(chisq.test(contingency_table)$expected < 5)) {
    # Performs chi-square test with simulated p-value for small expected frequencies
    chi_square_result <- chisq.test(contingency_table, simulate.p.value = TRUE, B = 1000)
  } else {
    # Performs chi-square test
    chi_square_result <- chisq.test(contingency_table)
  }
  
  # Checks if the p-value is significant
  if (chi_square_result$p.value >= 0.05) {
    significant_p_values <- FALSE
    break  # No need to continue checking if any p-value is not significant
  }
}

# Outputs statement based on the result
if (significant_p_values) {
  cat("All p-values are significant (less than 0.05)\nThere is sufficient evidence to conclude all associations between each latent class and morbidity are significant")
} else {
  cat("Not all p-values are significant (greater than or equal to 0.05)\n")
}
```

Run chi-square test to assess preferred homogeneity amongst classes
```{r}
latent_class_memberships <- lca_out$predclass  # Latent class memberships

# Creates a vector to track significance
significant_p_values <- logical(length(comorbidity_columns))

# Loops through each latent class
for (latent_class_number in unique(latent_class_memberships)) {
  subset_data <- final_table[latent_class_memberships == latent_class_number, ]
  
  # Loops through each variable of interest (e.g., comorbidity_columns)
  for (i in seq_along(comorbidity_columns)) {
    variable <- comorbidity_columns[i]
    chi_square_result <- chisq.test(table(subset_data[[variable]]))
    
    # Track the significance of p-values
    significant_p_values[i] <- chi_square_result$p.value < 0.05
  }
}

# Summarizes based on significance of p-values
if (all(significant_p_values)) {
  cat("All p-values are significant (less than 0.05)\nThere is sufficient evidence to conclude all occurences of morbidity groupings amongst each latent class is significant\n")
} else {
  cat("Not all p-values are significant (greater than or equal to 0.05)\n")
}
```

Run chi-square test for admission_type_encoded (categorical variable)
```{r}
admission_type_encoded <- final_table$admission_type_encoded  # Assuming it's a factor or character

# Performs chi-square test for admission_type_encoded and latent_class_memberships
chi_square_result <- chisq.test(table(admission_type_encoded, latent_class_memberships))

# Checks if the p-value is significant
if (chi_square_result$p.value < 0.05) {
  cat("The p-value is significant (less than 0.05)\n")
} else {
  cat("All p-values are significant (less than 0.05)\nThere is sufficient evidence to conclude all associations between each latent class and admission type are significant")
}


```

Run ANOVA test for age (quantitative variable)
Save LCA results into csv
```{r}
latent_class_memberships <- as.numeric(as.character(lca_out$predclass))  # Convert to numeric

# Creates a new data frame with numeric variables
data_for_anova <- data.frame(age = as.numeric(final_table$age), 
                             latent_class = latent_class_memberships)

# Performs ANOVA for age across latent classes
anova_result <- aov(formula = age ~ latent_class, data = data_for_anova)

# Extracts p-value from ANOVA results
p_value_anova <- summary(anova_result)[[1]]$"Pr(>F)"[1]

# Checks if the p-value is significant
if (p_value_anova < 0.05) {
  cat("The p-value from ANOVA is significant (less than 0.05)\n")
} else {
  cat("The p-value from ANOVA is not significant (greater than or equal to 0.05)\n")
}

```

Save LCA results into CSVs (each subgroup is saved into a separate csv)
For instructions on how to match our 7 subgroups to the 6 subgroups from the study, please check 04_morbidity_composition.ipynb on the github repository.
```{r}
final_table$subgroup <- apply(lca_out$posterior, 1, which.max) 
subgroup_counts <- table(final_table$subgroup)

output_folder <- "../processed_data/"

for (i in 1:7) {
  subgroup_data <- final_table[final_table$subgroup == i, ] 
  write.csv(subgroup_data, file.path(output_folder, sprintf("subgroup_%d.csv", i)), row.names = FALSE)
}
```


